---
title: "mccarthy_exploratory"
format: html
editor: visual
---

# Cormac McCarthy EDA

------------------------------------------------------------------------

```{r}
# packages
library(pdftools)
library(tidyverse)
library(dplyr)
library(purrr)
library(stringr)
library(udpipe)
library(epubr)
library(zoo)
# latest version of syuzhet package
# devtools::install_github("mjockers/syuzhet")
library(syuzhet)
```

## How do I read in a pdf to R?

```{r}
theroad_url <- "/Users/jonmichaelstroh/mccarthy_oeuvre_analysis/Books/theroad.pdf"

theroad_pdf <- pdf_text(theroad_url)

theroad_wrapped <- strwrap(theroad_pdf, width = 70)

# remove title, dedication, and "The End"
theroad_wrapped <- theroad_wrapped[5:4524]
```

## Explore UDPipe

```{r}
ud_model <- udpipe_download_model(language = "english")
ud_model <- udpipe_load_model(ud_model$file_model)

ud_pipe_model_road <- udpipe_annotate(ud_model, x = theroad_pdf, trace = FALSE)
theroad_word_frequency <- data.frame(table(as.data.frame(ud_pipe_model_road)$upos))

# clean the road word frequency table a tad
theroad_word_frequency <- theroad_word_frequency |>
  rename(word_type = Var1, word_frequency = Freq)

abbreviations <- c(ADJ = "adjective", ADP = "adposition", ADV = "adverb",
                   AUX = "auxiliary", CCONJ = "coordinating conjunction",
                   DET = "determiner", INTJ = "interjection", NOUN = "noun",
                   NUM = "numeral", PART = "particle", PRON = "pronoun",
                   PROPN = "proper noun", PUNCT = "punctuation",
                   SCONJ = "subordinating conjunction", SYM = "symbol",
                   VERB = "verb", X = "other")

theroad_word_frequency <- theroad_word_frequency |>
  mutate(word_type = recode(word_type, !!!abbreviations))
```

## Quickly, Try to Count Words

-   According to [wordcounters.com](https://wordcounters.com/word-count/literary-books/), "The Road" has 58,702 words

-   When I put into a google doc and removed all the line breaks and numbers, I got 58,740 words

```{r}
theroad_word_frequency |>
  filter(!word_type %in% c("punctuation", "numeral", "symbol")) |>
  summarise(total_words = sum(word_frequency))
```

-   Interestingly, we have categorized over 1000 extra words, something to monitor

-   Do certain words get double categorized?

-   Or, since Cormac McCarthy uses uncommon or created compound words such as "bloodcults" or "bottomlands" are they being split and then counted twice?

## Visualize Word Frequency in "The Road"

```{r}
ggplot(theroad_word_frequency, aes(x = reorder(word_type, -word_frequency), y = word_frequency)) +
  geom_bar(stat = "identity", fill = "gray", color = "black") +  
  labs(title = "Word Type And Frequency in 'The Road'",
       x = NULL,
       y = NULL,
       caption = "Data from the novel by Cormac McCarthy") +
  theme_minimal() +  
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(size = 14, face = "bold"),
        plot.caption = element_text(size = 8),
        plot.background = element_rect(fill = "white"),
        panel.grid.major = element_blank(),  
        panel.grid.minor = element_blank()) 
```

## Sentiment Analysis for "The Road"

-   Interested in using the Syuzhet package which reveals emotional shifts in the language of a story

-   Inspiration for this thinking came from the late Kurt Vonnegut on the Shape of Stories (I watched in my Intro to Creative Writing, inspiring writers to this day)

## How do I preprocess the data for Sentiment Analysis?

-   Julia Silge in [an analysis](https://juliasilge.com/blog/if-i-loved-nlp-less/) of Jane Austen's novels packages the text into 10 lines of about \~ 70 characters each for analysis

```{r}
# package into 700 character chunks with no words split in half
theroad_packaged <- tibble(text = theroad_wrapped) |>
  mutate(index = row_number(),                 
         line = ceiling(index / 10)) |>
  group_by(line) %>%
  summarize(text = str_c(text, collapse = " "))
```

Three sentiment lexicons:

-   `afinn` from [Finn Årup Nielsen](http://www2.imm.dtu.dk/pubdb/views/publication_details.php?id=6010)

-   `bing` from [Bing Liu and collaborators](https://www.cs.uic.edu/~liub/FBS/sentiment-analysis.html)

-   `nrc` from [Saif Mohammad and Peter Turney](http://saifmohammad.com/WebPages/NRC-Emotion-Lexicon.htm)

```{r}
# function to extract multiple sentiments from Lincoln Mullen
get_multi_sentiment <- function(sentences) {
  list(bing  = get_sentiment(sentences, method = "bing"),
       afinn = get_sentiment(sentences, method = "afinn"),
       nrc   = get_sentiment(sentences, method = "nrc"))
}
```

```{r}
# get sentiments
theroad_sentiments <- theroad_packaged |>
  mutate(sentiments = map(text, get_multi_sentiment)) |>
  unnest_wider(c(sentiments))
```

## Sentiment in 'The Road'

```{r}
theroad_sentiments %>%
  ggplot(aes(x = line, y = bing)) +
  geom_bar(stat = "identity") +
  annotate("label", x = 238.5, y = 5, label = "They discover the bunker filled with supplies", hjust = 0.5, label.size = 0, size = 3, color = "#2b2b2b") +
  annotate("segment", x = 238.5, xend = 238.5, y = 4.6, yend = 3.2, arrow = arrow(length = unit(0.02, "npc")), color = "#2b2b2b") +
  annotate("label", x = 173, y = 3, label = "They find the cellar\nof imprisoned people", hjust = 0.5, label.size = 0, size = 3, color = "#2b2b2b") +
  annotate("segment", x = 177.5, xend = 177.5, y = 2, yend = 0.5, arrow = arrow(length = unit(0.02, "npc")), color = "#2b2b2b") +
  theme_minimal() +
  labs(
    title = expression(paste("Sentiment in ", italic("The Road"), " by Cormac McCarthy")),
    y = "Sentiment",
    caption = "Bleak Novel LMAO"
  ) +
  theme(
    plot.title = element_text(size = 14, face = "bold", color = "#2b2b2b"),
    plot.caption = element_text(color = "#2b2b2b"),
    axis.title.y = element_text(color = "#2b2b2b"),
    axis.text.y = element_text(color = "#2b2b2b"), 
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  )

```

### Is the small rise in the center when they find the bunker filled with food?

-   Yes!

```{r}
# just to check which part of the novel for different sentiments
# delete later
combined_text <- theroad_sentiments |>
  filter(line >= 170 & line <= 185) |>
  summarise(full_text = str_c(text, collapse = " ")) |>
  pull(full_text)

text_chunks <- strsplit(combined_text, "\\s+")[[1]]
chunked_list <- split(text_chunks, ceiling(seq_along(text_chunks)/100))

# Print each chunk on a new line for readability
sapply(chunked_list, function(chunk) cat(paste(chunk, collapse=" "), "\n\n"))

```

## Let's try some more McCarthy novels...and maybe a few Atwood novels for comparison

```{r}
# download and process additional novels
# might have to read in epubs

alltheprettyhorses_pdf <- epub("/Users/jonmichaelstroh/mccarthy_oeuvre_analysis/Books/alltheprettyhorses.epub")[[10]][[1]] |>
  select(text) |>
  summarise(full_text = paste(text, collapse = " ")) |>
  substring(196)

prettyhorses_wrapped <- strwrap(alltheprettyhorses_pdf, 700)

prettyhorses_packaged <- tibble(prettyhorses_wrapped) |>
  rename(text = prettyhorses_wrapped) |>
  mutate(line = row_number())
```

```{r}
# pretty horses sentiment
prettyhorses_sentiments <- prettyhorses_packaged |>
  mutate(sentiments = map(text, get_multi_sentiment)) |>
  unnest_wider(c(sentiments))
```

-   Must note: Cormac writes in Spanish (mostly dialogue) occasionally and I'm unsure how the sentiment analysis would pick that up. Could I translate the entire text to English?

## Sentiment in 'All The Pretty Horses'

-   to use to label plot points:

```{=html}
<!-- -->
```
-   annotate("label", x = 238.5, y = 5, label = "They discover the bunker filled with supplies", hjust = 0.5, label.size = 0, size = 3, color = "#2b2b2b") + annotate("segment", x = 238.5, xend = 238.5, y = 4.6, yend = 3.2, arrow = arrow(length = unit(0.02, "npc")), color = "#2b2b2b") + annotate("label", x = 173, y = 3, label = "They find the cellar\nof imprisoned people", hjust = 0.5, label.size = 0, size = 3, color = "#2b2b2b") + annotate("segment", x = 177.5, xend = 177.5, y = 2, yend = 0.5, arrow = arrow(length = unit(0.02, "npc")), color = "#2b2b2b") +

```{r}
prettyhorses_sentiments %>%
  ggplot(aes(x = line, y = bing)) +
  geom_bar(stat = "identity", aes(fill = bing >= 0), width = 1) + 
  scale_fill_manual(values = c("FALSE" = "#36939b", "TRUE" = "goldenrod4")) +
  theme_minimal() +
  labs(
    title = expression(paste("Sentiment in ", italic("All The Pretty Horses"), " by Cormac McCarthy")),
    y = "Sentiment",
    caption = "More of a wild ride. Eh, see what I did there"
  ) +
  theme(
    plot.title = element_text(size = 14, face = "bold", color = "tan4"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.title.y = element_text(color = "tan4"),
    axis.text.y = element_text(color = "tan4"),
    plot.caption = element_text(color = "tan4"),
    legend.position = "none"
  )
```

## I want to try some smoother lines...let's see if anything looks good...

```{r}
# w geom_smooth()

prettyhorses_sentiments %>%
  ggplot(aes(x = line, y = bing)) +
  geom_smooth(color = "#36939b", method = "loess", span = 0.2, se = FALSE) + 
  theme_minimal() +
  labs(
    title = expression(paste("Sentiment in ", italic("All The Pretty Horses"), " by Cormac McCarthy")),
    y = "Sentiment",
    caption = "YAY for happy endings!"
  ) +
  theme(
    plot.title = element_text(size = 14, face = "bold", color = "tan4"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.title.y = element_text(color = "tan4"),
    axis.text.y = element_text(color = "tan4"),
    plot.caption = element_text(color = "tan4"),
    legend.position = "none"
  )
```

```{r}
# rolling average

prettyhorses_sentiments |>
  mutate(rolling_avg = rollmean(bing, k = 10, fill = NA, align = 'right')) |>
  filter(!is.na(rolling_avg)) |>
  ggplot(aes(x = line, y = rolling_avg)) +
  geom_line(color = "#36939b", linewidth = 1) +
  theme_minimal() +
  labs(
    title = expression(paste("Sentiment in ", italic("All The Pretty Horses"), " by Cormac McCarthy")),
    y = "Sentiment",
    caption = "More of a wild ride. Eh, see what I did there"
  ) +
  theme(
    plot.title = element_text(size = 14, face = "bold", color = "tan4"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.title.y = element_text(color = "tan4"),
    axis.text.y = element_text(color = "tan4"),
    plot.caption = element_text(color = "tan4"),
    legend.position = "none"
  )
```

## Load in all the rest of Cormac's Novels...

```{r}
# the orchard keeper 65


```